---
title: "Seurat Projection Score on PBMC3K"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)

set.seed(42)

pkg_root <- normalizePath(file.path(getwd(), ".."), mustWork = FALSE)
if (!file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  pkg_root <- normalizePath(getwd(), mustWork = TRUE)
}

pbmc3k_path <- Sys.getenv(
  "EMBOLCALL_PBMC3K_RDS",
  unset = "/Users/xpechuanjorge/Documents/Codex/2026-04-29/make-a-plan-to-take-the/examples/pbmc3k.rds"
)
```

# Goal

This notebook tests `SeuratProjectionScore()` on the local PBMC3K Seurat object.
It sweeps several variance-filtering thresholds up to 5,000 features, always
retaining the most variable features first, using scaled expression values.

# Load Packages

```{r libraries}
library(Seurat)
library(SeuratObject)
library(ggplot2)

devtools::load_all(pkg_root, quiet = TRUE)
```

# Load PBMC3K

```{r load-pbmc3k}
stopifnot(file.exists(pbmc3k_path))

pbmc <- readRDS(pbmc3k_path)
DefaultAssay(pbmc) <- "RNA"

pbmc
dim(pbmc)
length(VariableFeatures(pbmc))
head(VariableFeatures(pbmc), 10)
```

# Prepare Scaled Features

The object already has 5,000 variable features. Here we scale exactly those
features and use the scaled layer for projection scoring. Each threshold below
keeps the top `n` most variable features from this ranked list.

```{r scale-features}
max_features <- 5000
feature_counts <- c(500, 1000, 1500, 2000, 3000, 4000, 5000)
nboot_smoke <- 1

if (length(VariableFeatures(pbmc)) < max_features) {
  pbmc <- FindVariableFeatures(
    pbmc,
    assay = "RNA",
    nfeatures = max_features,
    verbose = FALSE
  )
}

top_features_5000 <- head(VariableFeatures(pbmc), max_features)
stopifnot(length(top_features_5000) == max_features)

pbmc <- ScaleData(
  pbmc,
  assay = "RNA",
  features = top_features_5000,
  verbose = FALSE
)

scaled_matrix <- GetAssayData(pbmc, assay = "RNA", layer = "scale.data")
dim(scaled_matrix)
stopifnot(nrow(scaled_matrix) >= max_features)
```

# Projection Score Sweep

The bootstrap count is intentionally small here so the notebook acts as a smoke
test. Increase `nboot_smoke` for final analysis.

```{r score-sweep}
projection_score_summary <- do.call(
  rbind,
  lapply(feature_counts, function(n_features) {
    features <- head(top_features_5000, n_features)

    set.seed(42 + n_features)
    elapsed <- system.time({
      score <- SeuratProjectionScore(
        pbmc,
        assay = "RNA",
        layer = "scale.data",
        features = features,
        use.variable.features = FALSE,
        NPCs = 3,
        nboot = nboot_smoke
      )
    })

    data.frame(
      variance_threshold = n_features,
      n_features = n_features,
      threshold_label = paste0(n_features, " most variable features"),
      reference = ifelse(n_features == 2000, "Seurat 2,000 reference", "Threshold sweep"),
      layer = "scale.data",
      NPCs = 3,
      nboot = nboot_smoke,
      projection_score = score,
      elapsed_seconds = unname(elapsed[["elapsed"]])
    )
  })
)

projection_score_summary
stopifnot(
  nrow(projection_score_summary) == length(feature_counts),
  max(projection_score_summary$n_features) == max_features,
  all(is.finite(projection_score_summary$projection_score))
)
```

# Plot

```{r projection-score-plot, fig.width=8, fig.height=4.8}
ggplot(
  projection_score_summary,
  aes(
    x = n_features,
    y = projection_score,
    color = reference
  )
) +
  geom_hline(yintercept = 0, linewidth = 0.35, color = "grey45") +
  geom_vline(xintercept = 2000, linewidth = 0.35, linetype = "dashed", color = "grey35") +
  geom_line(color = "#2F5D62", linewidth = 0.85) +
  geom_point(size = 3.4) +
  geom_text(
    aes(label = sprintf("%.4f", projection_score)),
    vjust = ifelse(projection_score_summary$projection_score >= 0, -0.75, 1.6),
    size = 3.3,
    color = "grey20"
  ) +
  annotate(
    "text",
    x = 2000,
    y = max(projection_score_summary$projection_score),
    label = "2,000 reference",
    angle = 90,
    vjust = -0.45,
    size = 3.2,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = feature_counts,
    labels = scales::comma
  ) +
  scale_color_manual(
    values = c("Seurat 2,000 reference" = "#E15759", "Threshold sweep" = "#2F5D62")
  ) +
  labs(
    title = "Projection Score Across PBMC3K Variance Thresholds",
    subtitle = "Each point keeps the top variable features up to the threshold; maximum threshold is 5,000",
    x = "Most variable features retained",
    y = "Projection score",
    color = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold")
  )
```
